---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# sdf

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/sdf)](https://CRAN.R-project.org/package=sdf)
<!-- badges: end -->

The goal of `{sdf}` is to provide an extensible spatial data frame. It is inspired by Rust traits where if you implement the correct generics, you get a ton of sweet benefits. 

By implementing the required generics your geometry type can become an `sdf` (spatial data frame).

## Installation

You can install the development version of sdf like so:

``` r
remotes::install_github("josiahparry/sdf")
```

## Example

```{r}
library(sdf)

data(guerry, package = "sfdep")

# using sf
sdf_sf <- as_sdf(guerry)

sdf_sf

# using rust
sdf_rs <- as_sdf(
  dplyr::mutate(guerry, geometry = rsgeo::as_rsgeom(geometry))
)

sdf_rs
```


```{r}
library(dplyr, warn.conflicts = FALSE)

sdf_sf |> 
  group_by(region) |> 
  summarise(total_crime = sum(crime_pers))

sdf_rs |> 
  group_by(region) |> 
  summarise(total_crime = sum(crime_pers))
```



## Required method implementations

As of now presently only 2 generics are required to get basic functionality

- `bounding_box()`
- `combine_geometry()`

... 

additional generics will be added here
